Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-10-24 notebooks/expected_and_actual_counts_ratio.Rmd
Add code to describe data set

Discuss purpose of notebook and steps to achieve that goal. Workflow to produce a volcano plot between any two clades.
List steps to do analysis (e.g., how to structure the data).
Do some sanity checks.

@@

Describe the entire dataset. How many clades? How many subsets? How many nucleotide sites? How many mutations to exclude?

```{r describe_data}

@@

```{r pseudocount}
# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count+0.5)/(filtered_SARS_mutation$expected_count+0.5)
```

@@

Insert information about the plot. Plot to visualize data.

```{r stacked_bar, echo=FALSE}
# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade",y="Normalized Mutation Count")
```

Insert information about the plot. Plot to visualize data.

```{r nucleotide_site, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=actual_count/expected_count, x=nt_site))+
  geom_point()+
  theme_bw()+
  labs(x="",y="")
```

Insert information about the plot. Plot to visualize data.

```{r INSERT, echo=FALSE}
#ggplot(filtered_SARS_mutation, aes(fill=INSERT, y=INSERT, x=INSERT))+
  #geom_tile()+
  #theme_bw()+
  #labs(x="",y="")
```

@@

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

@@

# Save and load variable because calculating p-values takes a long time.
'save(p_values,file="fisher_pvalues.Rda")'
load("fisher_pvalues.Rda")

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]),"observed_expected"],fisher_data[which(fisher_data$nt_mutation_clade==c[2]),"observed_expected"])})
```

>>>>

[INSERT] Discuss purpose of notebook and steps to achieve that goal.

@@

Describe the entire data set.

```{r describe_data}
# How many rows? How many columns?
dim(SARS_mutation)

# How many phylogenetic clades?
length(unique(SARS_mutation$clade))

# How many rows for each subset? Data from all countries, only England or only the United States.
data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA"),]))

# How many nucleotide sites?
length(unique(SARS_mutation$nt_site))

# How many mutations to exclude for each subset?
data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all" & SARS_mutation$exclude=="TRUE"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England" & SARS_mutation$exclude=="TRUE"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA" & SARS_mutation$exclude=="TRUE"),]))
```

@@

```{r pseudocount}
# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + 0.5)/(filtered_SARS_mutation$expected_count + 0.5)
```

@@

[INSERT] information about the plot. Plot to visualize data.

```{r stacked_bar, echo=FALSE}
# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

[INSERT] information about the plot. Plot to visualize data.

```{r nucleotide_site, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=actual_count/expected_count, x=nt_site))+
  geom_point()+
  theme_bw()+
  labs(x="", y="")
```

[INSERT] information about the plot. Plot to visualize data.

```{r INSERT, echo=FALSE}
#ggplot(filtered_SARS_mutation, aes(fill=INSERT, y=INSERT, x=INSERT))+
  #geom_tile()+
  #theme_bw()+
  #labs(x="", y="")
```

@@

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind", replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

@@

# Save and load variable because calculating p-values takes a long time.
'save(p_values, file="fisher_pvalues.Rda")'
load("fisher_pvalues.Rda")

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]), "observed_expected"], fisher_data[which(fisher_data$nt_mutation_clade==c[2]), "observed_expected"])})
```
